MLH1 Results Outline

### 2020-04-07 doc/Results_v2.rmd
adding Bret's re-writes and removing commented out sections

# Evolutionary Framework

@@

## Within Mouse Variance in CO Count per Cell

@@

@@

@@

<!--
- This is motivation to investigate more traits at the chromosome level
1.	Basic stats - bivalent data numbers 
2.	Rang of co per chrm, mostly 1co 2co
3.	High rec strains have evolved a higher proportion of 2CO
For males Chi squared test, p = low group,  p = MSM and PWD,  SKIVE is intermediate)
MSM 60% PWD 50% SKIVE 30% 20 - 10% remaining (low) strains
4.	explain why this complicates further comparisons (chromosome size effect * chromosome cl.ass)
-->

@@

<!-- breif background on longer female SC  (mouse, human) Gruhn-->

@@

<!--
In linear models (M2 and M3) for **mean short bivalent length**, sex is the most significant coefficient (M2 glm; p =`r summ.Q1.glm_short.SC_M2$coefficients[,4][8]` and M3 glm; p = `r summ.Q1.glm_short.SC_M3$coefficients[,4][8]`) with slight significant strain and interaction effects (glm; p=`r summ.Q1.glm_short.SC_M2$coefficients[,4][7]`, p =`r summ.Q1.glm_short.SC_M2$coefficients[,4][14]` respectively).

The following effects had significant p values for the **total SC** from the above model (for the linear models M2 and M3), subspecies _musculus_ (glm M2; p = `r sum.Q1.glm_totSC_M2$coefficients[,4][2]`), subspecies _molossinus_ (glm M2; p =`r sum.Q1.glm_totSC_M2$coefficients[,4][3]`), sex male (glm M2; p =`r sum.Q1.glm_totSC_M2$coefficients[,4][8]`), strain lew p =`r sum.Q1.glm_totSC_M2$coefficients[,4][5]`, strain g  p =`r sum.Q1.glm_totSC_M2$coefficients[,4][4]`,

For M3 in the mean total SC lengths there were many strain effects are significant, but the largest are sex (glm M3; p = `r sum.Q1.glm_totSC_M3$coefficients[,4][8]`)
but also some strain effects, G strain (glm M3; p = `r sum.Q1.glm_totSC_M3$coefficients[,4][2]`), LEW strain (glm M3; p = `r sum.Q1.glm_totSC_M3$coefficients[,4][3]`), PWD strain (glm M3; p = `r sum.Q1.glm_totSC_M3$coefficients[,4][4]`). (this means...). Supporting the strong sex difference and also indicating that SC length has evolved across strains within this panel, when chromatin condensation is summarized at this level.

(subsp*sex) mol.male  p =`r sum.Q1.glm_totSC_M2$coefficients[,4][10]`, sex.g  p =`r sum.Q1.glm_totSC_M2$coefficients[,4][11]`, lew.male p =`r sum.Q1.glm_totSC_M2$coefficients[,4][12]`
for the _mmdomesticues^WSB^ _, _mmdomesticues^G^ _, and _mmdomesticues^LEW^ _
-->

@@

# Q1. 1CO Position Sex Differences

@@

<!-- 
Brief lit review for IFD / interference expectations  
**Petkov et al 2007,  CO interference underlies sex differences in RR**

"Here we show that in mice, this is because of a shorter genomic interference distance in females than in males, measured in Mb. However, the interference distance is the same in terms of bivalent length. We propose a model in which the interference distance in the two sexes reflects the compaction of chromosomes at the pachytene stage of meiosis."

Chrm1 genetic map, 
(other human and mice refs 
Tease n Hulten 2004: -- no difference between MLH1 foci in males and females

DeBoer 2006, 13 -- measured chrm 1 in males and females, both were 2.8 microns -->

<!--  set up -->

@@

<!--
The differences between the sexes is slightly different depending on if the males are in the high rec group. In the low rec strains (Dom + KAZ)...
(both males and females IFD distributions have lower and upper thresholds around 25% and 75%) respectively. however the weight of observations are slightly shifted across the sexes. There's an enrichment of lower IFD^norm^ in females relative to males, while the opposite is true for the higher threshold -- where males have an enrichment of higher/long IFD^norm^ compared to females.

In the three high rec strains -- the distributions have a more striking difference of the lower threshold being much higher in males, aproximately 30% to 40% compared to 25% in females. Also while the female upper threshold is capped at 75%, in males (some strains have higher upper thresholds -- and it's more likely to observe IFD^norm^ above the threshold.

G - the upper and lower thresholds of females look pretty similar, the lower is closer to .2 than .25 -- female has more lower IFDs

WSB - higher upper threshold (female's have many below the 25% lower threshold, males have a few below 25% threshold)

LEW -- upper threshold looks very similar (.75) in males and females, much fewer low threshold obs in males,

KAZ  -- female upper threshold around .75, lower threshold around .25 the distribution is quite uniform and spread out.  
Male pattern; upper threshold slightly higher than .75 (.8) and lower threshold seems to extend beyond .25 -- but maybe low sampling contributes to the spread? (the spread in IFD values seems much larger for KAZ males compared to males from other strains.)

PWD -- female lower threshold extends much lower .25 (like .15?),  upper threshold capped at .75
Male lower threshold very different (higher like .3) the male upper threshold is above .75

MSM -- female thresdholds, upper set around .75(to .65) lower at mostly .25, some cells(mice) have more .1 to .2 IFD observatsions.  
Male upper threshold it at .75 while lower is set to .3 or .4.

SKIVE -- female upper threshold is capped at .75, lower threshold is between .25 to .3 but low rate of outliers below .25.   Male lower threshold is moved up, .4 or .35 and upper threshold moved up aprx .8
-->

<!-- lmer/ glms  **M1. Mixed Model** $$mouse \ average \ IFD ~=~ subsp * sex + rand(strain) + \varepsilon $$

**M2. Linear Model** $$mouse \ average \ IFD ~=~ subsp * sex * strain + \varepsilon $$

**M3. Linear model**

$$mouse \ average \ IFD ~=~  sex * strain + \varepsilon $$
-->

@@

<!-- junk for Q1.IFD

  For the normalized values in both M2 and M3, sex is a significant effect, increasing 
with males nrm.IFD in males.
  SKIVE*male is the only other consistently significant effect, which also increases the nrm.IFD measure (p = male-SKIVE p =  ).
For the mixed models of IFDs, sex is a significant effect for both raw and nrmIFD. (IFD.PER), sex is most significant effect (LRT: p = `r Q1.IFD.PER_M1_lmer.results$sex.results`)

<For the raw measures, in M2, I think the 2 SKIVE effects mean that the female raw IFD is shorter than the male IFD.raw. The other effects are for PWD and SKIVE (larger raw IFD from intercept.)  I think the MSM and MOLF interaction effects were too far down the list to sop up any variance. For the M3, only SKIVE*male effect is close to significant>

<remove the chrm size effects -- or check out model with chromosome length
to rule out chromosome size effects, (espically in KAZ) I re-run the ttests using the short and long biv data sets. Our predictions for these are that longer chromsomes will show more of the 'typical male pattern' while short chromosomes often have a more uniform positing (hence tighter)>

Plotting the single bivalent data for the short and the long data set show in KAZ -- there are two observations near the lowerest IFD1.PER, (even lower than female). These come from 2 different males and I looked up the images to confirm that the measures are accurate (by the bivalent).

The above plots display the raw and normalized area which seperate two foci on the same chromosome. In the normalized plot, 9 observations fall above the length threshold of 150.  The female measures fall more uniformly across the range of (0 to 1).
-->

@@

@@

<!-- basic stats
`r length(OG.MLH1.merge.Skel$fileName)` total skelentonize measures were made from all of the MLH1 pachtyene cell images. X observations removed for having extra signal from other cells. `r length(unique( all.remove ) )` measures removed as potential outliers (by examining the category specific plots)<pattern for the single bivalent, pos correlation, The predicted pattern for SC length becomes more nuanced when the data are split up by Chromosome class. Mainly that 1CO are shorter in the high recombining strains then low recombining strains. This is likely due to the fact that more physically longer chromosomes have 1CO in the low recombining strains, which pushes the mean SC length up. Where as in the high recombining strains, physically longer chromosomes are more likely to be in the 2CO group.  This supports a general pattern of tighter clustering of SC lengths across chromosome classes in the high recombininggroup. They have a lower probability of 2COs below a certain SC length threshold.>  -->

@@

<!--
In the short bivalent data set, analyzed by M1, both subspecies are 
 (glm; p= _Musculus_ `r sum.glmQ2_short.biv_M1$coefficients[,4][2]` and _Molossinus_ p= `r sum.glmQ2_short.biv_M1$coefficients[,4][3]`). While the models for the reduced data set suggest that the subspecies with the high recombining strains have evolved longer SC lengths --
long bivalent set in M1 -- 
Musc `r sum.glmQ2_long.biv_M1$coefficients[,4][2]` 
MSM strain `r sum.glmQ2_long.biv_M1$coefficients[,4][7]`
<And the strain effect for G is significant (glm; G p =`r sum.glmQ2_short.biv_M1$coefficients[,4][4]`).>

**M3 -- tests within subspecies!!** SO the M3 model can be run within species specific data -- to test the predictions (based on the variation in mean MLH1 counts), (use all 3 metircs) **the within subspecies M3 models don't fully meet the predictions
for total.Sc : there are significant Dom strain effects: in musc kaz is significant negative, several other strains have higher total SC, (MOL in the intercept is significant).
short biv: Dom: Significant G effect, MUSC no significant -- strain effects
long.biv:  Dom: no strain effects, MUSC no significant strain effects, MOL - not enough observations
**For the long bivalent dataset, In M1 where the subsp are tested, only Musc is significant (glm; Musc p= `r sum.glmQ2_long.biv_M1$coefficients[,4][2]`). The significant strain effects in M1 are slight just MSM (glm; MSM p= `r sum.glmQ2_long.biv_M1$coefficients[,4][7]`).**

**For the long bivalent dataset, In M2 (limited to just the strain effects) a few strain effects are significant (glm; MSM p = `r sum.glmQ2_long.biv_M2$coefficients[,4][5]`, SKIVE `r sum.glmQ2_long.biv_M2$coefficients[,4][7]` and CZECH p = `r sum.glmQ2_long.biv_M2$coefficients[,4][9]`).**

**THERE IS a weak signal of the high recombiningstrains having longer SC (having evolved) longer SC (different chromatin compaction)  While the mouse mean total sc are significantly different between the two groups --- this difference --- translated to the reduced bivalent data sets (p = long  p = short)**   -->

@@

@@

<!--
The high rec musc strains, _musculus^PWD^_ and _musculus^SKIVE^_, do not have signifcantly different single foci landscapes from the other musc low rec strains (ttest; p = `r Q2.1COpos_mus.av_Musc_ttest$p.value`). While _molossinus^MOLF^_ is significantly more centrally positioned compared to _molossinus^MSM^_ (t.test; p = `r Q2.1COpos_mus.av_Mol_ttest$p.value`). 
is that _domesticus^WSB^_ has significantly more terminal positioning,
and _molossinus^MOLF^_ having significantly more medial position.
A general pattern of strain specific effects
we found in looking at the 1CO positions, is that _domesticus^WSB^_ has significantly more terminal positioning, while MOLF has significantly more medial position.-->

<!-- don't report GLMs - since not difference in the ttests, GLMs because there's no difference. long p = `r Rec.group_LONG.BivData_F1post_Ttest$p.value` short 1CO, male, p = `r Q2.1COpos.PER_short_ttest_HL$p.value`> 
toal data set, short and long biv. There's only a difference between MSM and MOKF
test ppredictions, Dom = NULL,  -->

@@

<!--
given that the t.test don't return significant pvalues so we are not motivated to follow up with linear models. Males from WSB and MOLF strains have 'strang'unique patterns for the single foci positions, which don't fit with predicted patterns from the genome-wide recombination rate variation
<I ran another model that used the single bivalents observations to predict the normalize foci position with fixed effects of chromosome length, subspecies, strian and the interaction effects.>

In the above chunk I tried making another glm's which uses the single bivalent measures and include chromosome size for the switching the model to include interaction (instead of additive), all of the pvalues get lower. .. the chromosome length has the most significant effect, LEW by chrm effect are significant

In the reduced model, chrm length is the most significant, strain SKIVE, MOLF, G and LEW are significant strain effects -->

<!-- Just hold these for reference
**For the M2 (subsp)**
There are differences for the dom strains: G strain(p = `r summary(Q2.glm_mus.av_1COpos_M1)$coefficients[,4][4]`).

LEW (slight) (p =`r summary(Q2.glm_mus.av_1COpos_M1)$coefficients[,4][5]`).  
But big effects from Mol (p = `r summary(Q2.glm_mus.av_1COpos_M1)$coefficients[,4][3]`).  
and MSM strain (p = `r summary(Q2.glm_mus.av_1COpos_M1)$coefficients[,4][7]`)  
**For M3 (all strains)**
strain G (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][2]`).
strain Lew (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][3]`).
strain PWD ( p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][4]`).
slight strain MSM  (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][5]`).
very sig strain MOLF (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][6]`).
strain SKIVE (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][7]`).
strain KAZ (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][8]`).
-->

@@

<!-- hide these linear model results for now
for the first model, 
chromosome length `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][2]`

subsp MOL `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][4]`  
the mol obj above isn't showing the reight value. 
strain G   `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][5]`  
strain LEW   `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][6]`  
Strain MSM `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][8]`  
chrm * LEW `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][14]`  
chrm *MSM  `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][16]`

M2
-chromosome length, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][2]`  
-strain G, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][3]`  
-strain LEW, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][4]`  
-strain MOLF, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][7]`  
-strain SKIVE, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][8]`  
-chrm length LEW, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][12]` -->

@@

@@

<!--
**Linear model M1 ** $$mouse \ average \ IFD ~=~ subsp * strain + \varepsilon $$

**Linear model M2 ** $$mouse \ average \ IFD ~=~ strain + \varepsilon $$ -->

@@

<!--
RAW IFD M1 results for M1  -- `r sum.Q2.M1glm_IFD.raw$coefficients[,4][6]` PWD, 
MSM `r sum.Q2.M1glm_IFD.raw$coefficients[,4][7]`
SKIVE `r sum.Q2.M1glm_IFD.raw$coefficients[,4][8]`
and M2
for PWD, `r Sum.Q2.M2glm_IFD.raw$coefficients[,4][4]`
MSM `r Sum.Q2.M2glm_IFD.raw$coefficients[,4][5]`
and SKIVE   `r Sum.Q2.M2glm_IFD.raw$coefficients[,4][7]`

>>>>

# Partitioning variation in recombination rate

To situate variation in recombination rate within an evolutionary framework, we
We applied fit a series of models including subspecies, strain, and sex, to to our dataset of mean MLH1 foci counts from 187 mice to fit the patterns of variation across sex, subspecies and strain into an evolutionary framework. We began with a full mixed model (M1, see Methods), which showed that strain (random effect p < someSmallNumber), sex (p = 3.9710^{-5}), subspecies (p= 1.7210^{-4}), and subspecies * sex (p = 3.110^{-5}) each significantly affect recombination rate. After a general linear model including all factors as fixed effects (M2) revealed only weak contributions of subspecies, we focused on additional models designed to illuminate the role of strain and sex. A general linear model with these two variables (M3) identified two strains with particularly strong effects on recombination rate: musculusMSM (p= 3.9910^{-6}) and domesticusG (p= 1.0410^{-6}). In addition, two strains exhibit strain-by-sex interactions:  molossinusMSM  (p = 1.2610^{-4}) and musculusPWD (p = 3.8610^{-4}). 
We next fit general linear models separately for 192 males and 144 females (M4). In the male dataset, three strains significantly affect recombination rate: molossinusMSM (p = 2.2310^{-14}; effect = X MLH1 foci), All of the effects within the mixed model M1 were highly significant (LRT: subsp p= 1.7210^{-4}, sex p = 3.9710^{-5}, subspecies * sex p = 3.110^{-5}) and random LTR; strain p= 0 ).
M1. Mixed Model 
mouse av. MLH1 foci per cell = subsp*sex+rand(strain)+??
M2. Linear Model
mouse av. MLH1 foci per cell = subsp*sex*strain+??
M3. Linear Model
mouse av. MLH1 foci per cell = sex*strain+??
M4. Linear Model
sex specific mouse av. MLH1 foci per cell = subsp*strain+??

The linear model results confirm the qualitative observations of sex-specific evolution and significant difference in the degree of sexual dimorphism in genome wide recombination rates. The subspecies fixed effects tested in M2, were either weak or not significant (glm; p=0.22 and p= 0.09 for M. m. musculus and M. m. molossinus respectively), indicating that most of the variance is due to strain effects. As such, we focus on results from M3. The two most significant strain effects were domesticusG and musculusMSM (p= 1.0410^{-6} and p= 3.9910^{-6} for domesticusG and molossinusMSM respectively). In addition to significant strain effects for molossinusMSM, (glm; p = 0.06 and p = 3.9910^{-6}) both musculusPWD and musculusMSM had significant strain by sex effects (glm; p = 3.8610^{-4} and p = 1.2610^{-4}) musculusPWD by male and musculusMSM by male. Combined, these results demonstrate that i) the significant sex-specific evolution in MSM and PWD and ii) the larger magnitude of heterochiasmy in domesticusG.
A linear model was run on the sex specific data sets (M4), with 192 male and 144 female mean MLH1 foci per cell. There were three significant strain effects in the male specific data; musculusPWD ((glm; p = 7.3710^{-10}; effect = X foci), and musculusSKIVE (glm; p = 0.01; effect = X foci). These three strains point to rapid evolution in recombination rate in spermatocytes; we subsequently refer to them collectively as "high-recombination" strains. Analysis of the female dataset points to four strains with significant effects on recombination rate:  and molossinusMSM (glm; p=2.2310^{-14} with effect sizes ranging from 5, 7, and 2 foci respectively.  Given these results, theses strains are classified as 'high recombining' strains and point to rapid evolution in the recombiantion rate for spermatocytes.
While the female specific linear model has many more had significant strain effects; domesticusG (glm; p = 2.510^{-6}), molossinusMSM (p = 6.2410^{-6}), domesticusLEW (glm; p = 0.01), and domesticusPWD musculusPWD (glm; p=0.02). Strain effect sizes in females are modest in magnitude (ranging from 1 to 4 foci) compared to those in males., domesticusMSM (glm; p=6.2410^{-6}), domesticusMOLF(glm; p=0.08), and domesticusKAZ(glm; p=0.1), the effect sizes have a smaller range of (4 to 1), indicating significant but small strain specific effects on mean MLH1 foci per cell.
Together, these results demonstrate heritable differences in the genome-wide recombination rate evolving in a highly sex-specific manner over short evolutionary timescales.

@@

## Within mouse variance in CO count per cell

Counting MLH1 foci in multiple oocytes for each female and multiple spermatocytes for each male allowed us to examine determinants of the within-mouse variance in recombination rate. To do this, we considered the same models as above, but replaced mean MLH1 foci count with within-mouse variance in MLH1 foci count as the dependent variable. Sex is the only variable that significantly affects recombination rate in both the mixed model (M1) (p < smallNumber) and general linear model (M2) (p = 2.310^{-4}). In general, females have almost twice as much variance in MLH1 foci per cell compared to males (Figure 1). Since estimates of within-mouse variance may be more susceptible to technical error from the staining protocol, we repeated the analyses using a subset of cells with higher quality scores (criteria?). The results are similar: sex is the strongest effect (p < smallNumber, M1; p = 2.310^{-4}, M2). When both quality-curated and full datasets are considered, strain does not significantly and consistently affect variance in MLH1 foci count per cell in either sex. These results suggest that within-mouse variance in recombination rate evolves independently of mean recombination rate.

@@

Evolution of genome-wide recombination rate is associated with evolution of double strand breaks

In an attempt to localize the male-specific evolution of crossover number to steps of the meiotic pathway, we counted foci from a marker for double strand breaks (DSBs), DMC1, in prophase spermatocytes. DMC1 foci were scored from a total of 76 leptotene-stage and 75 zygotene-stage spermatocytes from juvenile mice (12 to 18 days) from three low-recombination strains (musculusKAZ , domesticusWSB , and domesticusG) and two high-recombination strains (musculusPWD and molossinusMSM).
The high-recombination strains have significantly more DMC1 foci than the low-recombination strains in leptotene cells (t-test, p<someSmallNumber; one-way ANOVA, p = 0.00027). In contrast, the two strain groups do not differ in DMC1 foci counted in zygotene cells (later prophase)(t-test, p = 0.66; one-way-anova, p = 0.15).
After DSB formation, DSBs are repaired as either non-crossovers (NCO) or as crossovers (COs), with the vast majority being repaired as NCOs. Thus the ratio of CO:DSB is a partial indicator of the proportion of DSBs which are designated as COs. The ratios, calculated for DMC1 means from both stages, are not significantly different between the high and low strain groups (t-test, p = 0.94 and p = 0.11 for leptotene and zygotene ratios, respectively). This comparison raises the possibility that the evolution of crossover number is primarily due to processes that precede the crossover/non-crossover decision.

@@

Evolution of genome-wide recombination rate is reflected at the single chromosome level

To examine the connection between evolution of the genome-wide recombination rate and changes to the recombination landscape, we used an image analysis pipeline to measure properties of single bivalents (Peterson et al. 2019). This algorithm substantially speeds the accurate measurement of bivalents, but has the limitation that not all bivalents per cell can be isolated due to overlapping bivalents. In this dataset, isolation rates per cell range from 0.51 (molossinusMSM male) to 0.72 (musculusKAZ female). From the total set of cell images, 10,458 bivalent objects were isolated by the image analysis software. After a human curation step (Peterson et al. 2019), 9,829 single-bivalent observations remained. We assume that the isolation of bivalents within cells is unbiased. Given the large number of single-bivalent observations, we assume that each of the datasets are equally representative of general patterns.
An additional challenge of the MLH1 framework is that the identities of individual autosomes and the XX in females cannot be easily obtained (the male XY is distinct).

<figure>

Ninety-six percent of single bivalents in our pooled dataset (n = 34,982) have either one or two crossovers (Figure X). The proportion of one-crossover (1CO) to two-crossover (2CO) bivalents distinguishes the high vs. low recombining strains (Figure X). High-recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents: proportions of 2CO bivalents  are 0.33 in musculusSKIVE , 0.44 in musculusPWD , and 0.53 in molossinusMSM . Following patterns in the genome-wide recombination rate, male musculusPWD and male molossinusMSM have 2CO proportions that are more similar to each other than to strains from their own subspecies (chi-square tests; musculusPWD vs. musculusKAZ p = 3.1510^{-33}; molossinusMSM vs. molossinusMOLF p = 4.7210^{-13}). 
Next, we use this single-bivalent dataset to focus on aspects of the recombination landscape along chromosomes. We address two main questions. First, which traits are sexually dimorphic? Second, which traits differ between males from high-recombination vs. low-recombination strains?

@@

@@

Q1 Sex Differences in the Length of the Synaptonemal Complex

In many mammalian species, the synaptonemal complex (SC) is longer in females. Yet, the majority of these observations come from species with female-biased heterochiasmy. Our wider survey of recombination in house mice provides an opportunity to determine whether sex differences in chromatin compaction (SC length) are reversed when heterochiasmy is male-biased. In addition, if SC length is a strong determinant of the genome-wide recombination rate,  male musculusPWD and male molossinusMSM should have longer SC lengths than the other strains. 
To account for confounding effects of sex chromosomes from pooled samples of bivalents, we considered a reduced dataset including only bivalents with SC lengths below the 2nd quartile (for SC length) in each cell. This dataset included the four or five shortest bivalents  and excluded the X bivalent in oocytes. A total of 678 'short' bivalents were isolated from 103 oocytes and 37 spermatocytes. Although this smaller dataset has decreased power, it offers a more comparable set of single bivalents to compare between the sexes.

<figure>

Short bivalents are significantly longer in females than males in all strains (t-test; p < 0.05) except musculusSKIVE.  The female:male ratio ranges across strains from 1.15 (musculusMSM) to 1.49 (domesticusWSB). 
As an additional metric of chromatin compaction, we computed the total (summed) SC length of all bivalents for single cells, using the image analysis algorithm of Wang et al. (2019). Mouse means were calculated from cell-wide total SC lengths in 2,984 out of 3,680 cells with MLH1 counts (Figure X). Females have significantly longer total SC than males in each strain tested separately (t-test; p < 0.05). That females have longer SCs is also supported by mixed models and general linear models with covariates, which identify sex as the most consistently significant effect (p < 0.05). Additionally, there are some significant subspecies and strain effects (p < 0.05), indicating that SC length has evolved among strains and subspecies.

<figure>

In summary, two approaches for measuring and analyzing SC length indicate that females have longer SCs (chromosome axes), even in strains where males have more MLH1 foci per cell. Furthermore, joint consideration of MLH1 foci and total SC length suggests that males from high-recombination strains have less "space" in which to place their additional crossovers.

@@

@@

# Q1. 1CO Position Sex Differences
Q1. Sex Differences in the Positions of Single Crossovers

In most of the strains we surveyed, the majority of bivalents are observed to contain one crossover (focus). In all strains, the landscape across 1CO bivalents is significantly different in females and males. Normalized foci positions tend to be more central in females (mean focus position = X) and closer to the telomere in males (mean focus position = Y) (t-test; p = 2.9210^{-22}). Sex is also the most significant effect on focus position in a mixed model (M1: p =1.2610^{-25}) and in general linear models (M2: p =1.3310^{-7}; M3: p = 1.3310^{-7}). These sex differences in the placement of foci on 1CO bivalents follow a pattern observed across a variety of mammalian species (Sardell Kirkpatrick).

@@

Q1. Sex Differences in CO Interference (IFD)
Crossover interference is one of the major determinates of the recombination landscape. It generates a distribution of evenly spaced crossover along chromosomes instead of a random and more uniform distribution. To quantifiy how crossover interference differs between the typical recombination landscapes of males and females, we isolated 1360 and 1272 2CO bivalents in females and males respectively. Mean interfocal distances (IFD) were calculated from 42 female and 45 male mice. We examined both raw IFDraw and normalized by SC length (IFDnorm). The mean IFDraw measures of crossover interference as a mechanical force mediated through the SC. We compared the IFDraw to test the hypothesis of interference acting as a mechanical force transmitted through the SC. Previous measures found no sex difference in measures of IFDraw in mice (de Boer et al 2004). 
There is no strong signal of sex differences in raw mean inter-focal distances (IFDraw) across the full set of strains. A marginally significant difference between the sexes (t-test; p = 0.07) is driven by one strain (t-test without domesticusG; p = 0.27). The difference in mean IFDraw are slightly significant sexes from the full data set (t.test; p = 0.07). However domesticusG might be an outlier (by having a larger than average degree of sexual dimorphism). When the domesticusG observations are removed, the difference is not longer significant (t.test; p = 0.27). This result indicates there is no general pattern of sex differences in crossover interferencethat females and males exhibit a similar level of interference when it is measured in physical (SC) units. In contrast, males have significantly longer normalized mean inter-focal distances (IFDnorm) than females in seven out of eight strains (t-tests; p < 0.02), with only musculusKAZ showing no difference (t-test; p = 0.33). 
While the comparisons of the IFDnorm metrics can reveal more general recombination landscape patterns while controlling for the underlying differences in chromatin compaction and SC length. This metric is closer to measures of interference from linkage maps which are also removed from physical scales in that they measure frequencies of crossovers. Crossover interference is stronger in male specific linkage maps compared to female maps in a variety of species (ref).
The difference in mean IFDnorm are highly significant between sexes. The mean IFDnorm are significantly long in six of seven strains (MSM t.test; p-value = 0.01), musculusSKIVE p-value = 4.310^{-5}), musculusPWD p-value = 2.2810^{-4}), molossinusMSM p-value =0.01), domesticusLEW p-value = 3.5710^{-5}), domesticusG p-value = 0.05), and domesticusWSB (t.test; p-value =0.03) ). Only musculusKAZ did not have a difference in mean IFDnorm (t.test; p-value = 0.33).
In examining theExamination of IFDnorm distributions, we note indicates that  the female IFDnorm observations values are centered at approximately 50% and have show a slight an enrichment of short low (<25%) IFDnorm observationsvalues, whereas males are enriched for higher values. Comparitively, the IDFnorm distributions in males are enriched for longer IFDnorm.
The three sets of modelsMixed models and general linear models of IFDnorm support the result inference of stronger interference in males:, sex was theis the most significant significant effectvariable (LRT: p = 6.7410^{-14}). When interference is measured in physical SC units (IFDraw), the differences between sexes is low and only slightly significant (data not shown). 
In summary, controlling for differences in chromatin compaction (SC length) using IFDnorm indicates that interference is stronger in males, whereas consideration of IFDraw shows that the sexes exhibit a similar level of interference on the physical (SC) scale.
These results support the model of the physical measures of interference (in SC units) is conserved between sexes inwithin the same species.While the comparisons of the IFDnorm metrics can reveal more general recombination landscape patterns while controlling for the underlying differences in chromatin compaction and SC length. This metric is closer to measures of interference from linkage maps which are also removed from physical scales in that they measure frequencies of crossovers. Crossover interference is stronger in male specific linkage maps compared to female maps in a variety of species (ref).

Q1 Summary, consequences of sex differences in the recombination landscape
Our results confirm sex differences in recombination landscapes described in several other species and we note that these differences have distinct consequences on the potential patterns of genetic variation resulting from the distinct recombination landscapes.
The terminal position of single foci and the greater distance between two foci on the same chromosome will have a consequence of larger sections of linked sites segregating together in male gametes compared to female gametes (Veller).

Additional Determinants of Genome-wide Recombination Rate Evolution in Males
The next section is meant to focusNext, we used the contrast between high-recombination strains (musculusPWD , musculusSKIVE, and molossinusMSM) and low-recombination strains to identify features of the recombination landscape associated with evolutionary transitions in the genome-wide recombination rate. 
on the greater aspect of variation in mean MLH1 counts per cell the high recombining strains ( musculusPWD , musculusSKIVE, and molossinusMSM) to the low recombing strains. The main objective of this section is to test for significant correlations between features of the recombination landscape and the evolution of mean MLH1 foci per cell.
In comparing the male specific single bivalent based metrics the first creteria for analysis is a significant differences between the high and low recombining groups and the second step is testing for significant subspecies and strain effects in the same models which used mean MLH1 count per cell.

<equations>

**Linear model M1**

mouse average bivalentmetric = subsp*strain+??
Linear model M2 
mouse average bivalentmetric = strain+?? 
The literature provides Previous empirical work suggests basic predictions for the direction of the relationship between the genome wide recombination rate and two aspects of the recombination landscape.; SC length is expected to be positively associated with genome-wide recombination rate because X. Crossover interference is expected to be negatively associated with genome-wide recombination rate because X. (positive) and crossover interference (negative). Following this logic we predict i) no strain effects for SC length or crossover interference within domesticus, ii(1) musculusPWD will have greater SC length and weaker interference than musculusSKIVE, which in turn will have longer SC and weaker interference compared to the other musculus strains, and iii(2) molossinusMSM will have longer SC and weaker interference compared to molossinusMOLF, and 3) domesticus strains will have similar SC length and crossover interference..

@@

@@

@@

We compared mouse means of three SC-based metrics: total SC length, mean short bivalent length, and mean long bivalent length. The long bivalent data set was isolated by choosing bivalents in the X quartile of the SC length distribution within cells. The short bivalent dataset was isolated as described above.

figure

Confirming the basic predictions, there is a positive correlation between mean MLH1 foci per cell and total SC ([report spearman's rho and p-value]). Nevertheless, mean total SC only partially differentiates high-recombination and low-recombination strains (Figure X). While high-recombination strains have significantly more SC area in the total dataset (t-test; p = 0.01), separate tests by strain show that only within molossinus are high- and low-recombination strains significantly different (t-test; molossinus: p = 0.03; musculus: p = 0.87). Additionally, the means for the reduced (short and long) bivalent datasets are not significantly different between high-recombination and low-recombination strains (t-test; short: p = 0.88; long: p = 0.18).
In a general linear model with total SC as the dependent variable, two subspecies effects are significant (p = musculus 1.2410^{-6}, molossinus p = 10^{-6}). In general linear models with reduced bivalent means as dependent variables, several subspecies and strain effects reach significance (p < 0.05) but they are not consistent across models, indicating to some extent the chromatin compaction evolution is decoupled from evolution in mean MLH1 foci per cell.

<bret's version above>

@@

@@

@@

The normalized 1CO position is not significantly different between high-recombination and low-recombination strains for the total pooled data (t-test; p = 0.24) and also when examined within subspecies (t-test; p = 0.41 and p = 0.07 for musculus and molossinus, respectively). While there are significant strain effects for domesticusWSB and molossinusMOLF in a general linear model with normalized position as the dependent variable (Figure X), this evolution of the 1CO positioning is decoupled from the total genome-wide recombination rate.

@@

@@

@@

@@

Mouse averages for both IFDraw and IFDnorm are significantly longer in high-recombination strains (t-test; IFDnorm: p = 7.7410^{-7}; IFDraw: p = 8.7810^{-6}). This pattern is confirmed by separate comparisons within musculus (t-test; IFDnorm: p =2.0410^{-5}; IFDraw: p = 1.9410^{-4}) and within molossinus (IFDnorm: p= 0.17; IFDraw: p = 0.08). Similar results are seen with general linear models for both IFDraw and IFDnorm: only effects associated with high-recombination strains are significant (p < 0.05). That IFDraw and IFDnorm show similar patterns eliminates variation in SC lengths and bivalent sizes as primary explanations. We determined that the main difference in IFDnorm distributions between high-recombination and low-recombination strains is an enrichment of IFDnorm values under 30% in low-recombination strains. The frequency of IFDnorm values that fall below 30% ranges from 8.2% (domesticusG) to 16% (musculusKAZ) in low-recombination strains, whereas high-recombination strains all show such frequencies below 5% (0%, 1.3%, and 3.3% for musculusSKIVE, molossinusMSM, and musculusPWD, respectively).
In summary, the level of interference is a significant predictor of evolution in the genome-wide recombination rate, but SC length and crossover position on 1CO bivalents are not. However, the pattern is in the opposite direction to our prediction: high-recombination strains have stronger interference. At least on 2CO bivalents, crossovers are spaced further apart when the genome-wide recombination rate is higher.

@@

@@
